### 2015-04-30 internal_review_response.Rnw

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{indentfirst}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[final]{animate}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{array}
\usepackage{subfig} 
\usepackage[noae]{Sweave}
\usepackage{cleveref}
\usepackage[figureposition=bottom]{caption}
\usepackage{paralist}
\usepackage{acronym}
\usepackage{outlines}

%acronyms
% \acrodef{}{}

%knitr options
<<setup, cache = F, echo = F>>=
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, echo = F)
options(replace.assign=TRUE,width=90,digits=1)
@

\setlength{\parskip}{5mm}
\setlength{\parindent}{0in}

\begin{document}
\raggedright

% \title{}
% \author{}
% \maketitle

{\it Response to review from Dr. Jeff Hollister, ``SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries'', Beck, MW.}

Response to reviewer comments are provided in italics.  The original review is included as a pdf with markup comment. The original pages and line numbers are noted below.

Page 1: I describe SWMPr, an open-source R package, for use ...

{\it Changed.}

Page 2: growth

{\it Changed.}

Page 2: is limited

{\it Changed.}

Page 2: identifying the signal from the

{\it Changed.}

Page 2: This feels like you are questioning the use of this type of data? Also, try to transition to the next paragraph. Feels a bit abrupt.

{\it Modified as follows: `A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs [6, 7].  An important monitoring network for coastal environments in the United States is the National Estuarine Research Reserve System (NERRS). This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.'}

Page 2: both for

{\it Changed.}

Page 2: Why just abiotic? You should add a bit about this. Are the other data not available, just haven't done it yet? If the later, let me know, that'd be fun to work on!

{\it The description in the text is vague in this regard and I have modified for clarity: `Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the SWMP data and, consequently, the most challenging to evaluate.  '}

Page 3: space

{\it Added.}

Page 3: On CRAN now, correct? IF so update installation directions.

{\it Installation instructions were updated}

Page 3: Would bet a lot of readers won't have any idea what this means. Since the swmpr object is central to the package, maybe a section describing its structure and a line or two on S3 might be a good idea.

{\it Agreed.  This material was already in the text but not as a separate section.  I added a section describing the object structure, including the following text for clarity: `The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system, [16]), such that specific methods for a generic function are developed for the object class.}

Page 4: Why only 100? web service limit?

{\it Added footnote to the table that describes the limitation.}

Page 4: Same as above. If there are service limitations should outline those. Just saw this explained below!

{\it Same as above.}

Page 4: This is going to be a tough one to keep relevant. What's the reason for including it?

Page 4: Couldn't you build this functionality into the function itself? Perhaps next version... No need to include prior to publishing this.

Page 5: Might be more informative to show the results for each request... dplyr style?

Page 5: If you built in the loop to get around the limit would this not be necessary? This seems like a bandaid that will be less and less relevant over time.

Page 5: Unless it is necessary, I wouldn't even mention this.

Page 5: Would be nice to have the decompression as part of the function.

Page 5: What is the source of the duplicates? Does it make sense to just remove them? Are the data duplicated? Could you just flag these in the output as opposed to removing? If you did this, you could then let the user decide when to remove or not remove with rem\_reps().

Page 6: Re: my comment above about the object... Might make this a separate section. Include a table instead of prose to describe the object. Could closely follow the output of str(my\_swmpr\_obj).

Page 7: setstep.swmpr??

Page 7: Not sure how best to refer to these S3 methods. You do it both ways (with .smpr in the table) and without in the text. Could keep it that way if you add a line to table caption. I bring it up, because I think new R users would be confused by this (I was for awhile...)

Page 9: Does PLOS allow Boxes? Like Box 3. Example of qaqc() funciton usage... That would be a good way to show these.

Page 9: quote

Page 9: quote

Page 10: A section on package dependencies might be a good addition.

Page 10: Include an example of this in the code examples below.

Page 10: rewrite. ``a signal from noisy data'' ??

Page 13: or

Page 14: May not be obvious that this is 60 min / 15 min... And character string options are a great idea!

Page 16: I'm just being picky, but the name on this one is tripping me up. I think I'd prefer to see eco\_metab. But, I realize that may not be an easy change, so up to you!

Page 17: Do all of these functions need to be exported? Could they simply be internal functions that the users don't see? If so, you could shorten or eliminate this section. Also you could create a Viz category of functions instead of this one. In that you could talk about the plot capabilites separate from the analysis section, and the include the reserve\_map function there.

Page 17: Should this be map\_reserve???

Page 19: Yank this. Goes without saying.

Page 20: Some of the figures have code that directly preceeds it. Other figures (like Fig 8) it is a bit more difficult to link code to figure. Would be nice to provide a link to code for all the figures. This way a reader could download a single script to reproduce the figures.

Page 20: Are these examples necessary? Feel kinda tacked on and why have links as supplementary info and not part of the text? If you are going to keep these, I would flesh this out a bit more, add in some screen captures as figures and links directly in the text.

Page 21: Same comment as above. Be nice to have ready access to code that reproduces this table.

\end{document}

### 2015-04-30 internal_review_response.tex

\documentclass[letterpaper,12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{indentfirst}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[final]{animate}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{array}
\usepackage{subfig} 
\usepackage[noae]{Sweave}
\usepackage{cleveref}
\usepackage[figureposition=bottom]{caption}
\usepackage{paralist}
\usepackage{acronym}
\usepackage{outlines}

### 2015-04-30 swmpr_manu.Rnw
started revisions from jwh

% acronyms
\acrodef{CDMO}{Centralized Data Management Office}
\acrodef{DO}{dissolved oxygen}
\acrodef{NERRS}{National Estuarine Research Reserve System}
\acrodef{QAQC}{quality assurance/quality control}
\acrodef{SWMP}{System Wide Monitoring Program}

@@

\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example in the United States is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS}.  This program has provided two decades of continuous monitoring data at over 300 fixed stations in 28 estuaries.  \ac{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  We describe a new open-source software package, SWMPr, developed in program R for use with \ac{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.

@@

The development of low-cost, automated sensors that have the ability to collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.  These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.  Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \cite{NDBC15}.  Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.  However, the growing quantity of available information to address relevant questions has contributed to the notion of `big data' science where analysis limitations are often defined by compututational requirements and signal identification in the presence of noise rather than the availability of information.  Growing concerns over the use of adequate quality assurance and control methods, challenges for synthesis and interpretation, and increased emphasis on exploratory-based analytical techniques have characterized applications of data from automated monitoring programs \cite{Campbell13,Millie13}.

The \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}} is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.  The reserves represent different biogeographic regions and estuarine types that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 300 stations at the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied in evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.  \ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.

@@

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  The SWMPr package is developed for the continuous abiotic monitoring network which includes a majority of the data collected by \ac{SWMP}.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal\footnote{\url{http://cdmo.baruch.sc.edu/}}, where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.

@@

The SWMPr package was developed for use with the R ($\geq$v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be installed from GitHub\footnote{\url{https://github.com/fawda123/SWMPr}} by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

<<eval = F, message = F>>=
install.packages('devtools')
library(devtools)
install_github('fawda123/SWMPr')
library(SWMPr)
@
<<eval = T, echo = F, message = F, cache = F>>=
devtools::load_all('M:/docs/SWMPr')
@

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes a general approach for working with \ac{SWMP} data using each category of functions.

@@

descrips <- c(
  'Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter.  Maximum of 1000 records. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
  'Import site data from an independent server. These files have been downloaded from \\ac{CDMO} up to December 2014, processed using SWMPr functions, and uploaded to an Amazon server for quicker import into R.',
  'Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
  )

@@

latex(
  to_tab[, 'Description', drop = F],
  file = '',
  caption = 'Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).',
  rowlabel = 'Function',
  colheads = 'Description',
  rowname = to_tab$Functions,
  caption.loc = 'top',
  col.just=c("p{3.5in}"), 
  label = 'tab:retrieve', 
  table.env = FALSE
  )
@

@@

All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.  The data include a \texttt{datetimestamp} column in the appropriate timezone for a station.  Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg \ac{QAQC} columns if returned from the initial data request.  The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).  Attributes of a \texttt{swmpr} object can be viewed as follows.

@@

The \texttt{swmpr} object class was created for use with specific methods following the S3 object definition approach \cite{Wickham14}.  A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods also apply to \texttt{swmpr} objects.  Available methods for the \texttt{swmpr} class are described below and can also be viewed:

@@

The ability of management and research programs to address critical environmental issues is highly dependent on the quality of data used to inform decision making.  Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessments of site-specific characteristics and more informed decisions to manage environmental resources.  The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 300 stations within the 28 estuaries of the \acl{NERRS}.  This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.  Although the data provided by \ac{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \ac{NERRS} researchers and staff to more effectively evaluate \ac{SWMP} data could be greatly improved using the SWMPr package.

>>>>

% acronyms
\acrodef{CDMO}{Centralized Data Management Office}
\acrodef{CRAN}{Comprehensive R Archive Network}
\acrodef{DO}{dissolved oxygen}
\acrodef{NERRS}{National Estuarine Research Reserve System}
\acrodef{QAQC}{quality assurance/quality control}
\acrodef{SWMP}{System Wide Monitoring Program}

@@

\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example in the United States is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS}.  This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.  \ac{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with \ac{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.

@@

The development of low-cost, automated sensors that collect data in near real-time has enabled a proliferation of standardized environmental monitoring programs \cite{Glasgow04,Fries08}.  These programs provide access to invaluable sources of data that can be used to address a variety of research and management objectives.  Applications from automated remote sensors are numerous for aquatic environments with notable examples including prediction of harmful algal blooms and toxicants in freshwater systems \cite{Reed10}, development of a hydrometeorological monitoring network to support flash flood warning programs \cite{HADS15}, and a national marine buoy network covering large portions of the open ocean and coastal zones of the United States \cite{NDBC15}.  Automated remote monitoring programs offer several advantages over traditional site-specific, field-based methods including streamlining of data acquisition, minimizing human error, and reducing the overall cost of the collection process \cite{Glasgow04}.  However, the increasing quantity of available information to address relevant questions has contributed to the growth of `big data' where analyses are limited by compututational requirements and identifying the signal from the noise rather than the availability of information. A greater focus on synthesis, exploratory-based analytical techniques, and interpretation have characterized the use of data from automated monitoring programs \cite{Campbell13,Millie13}.

An important monitoring network for coastal environments in the United States is the \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}}.  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 140 stations at the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.  \ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.

@@

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \ac{SWMP} data and, consequently, the most challenging to evaluate.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal\footnote{\url{http://cdmo.baruch.sc.edu/}}, where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.

@@

The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be downloaded from the \ac{CRAN}\footnote{\url{http://cran.r-project.org/web/packages/SWMPr/index.html} for the \ac{CRAN} release, the development version can be installed  from \url{https://github.com/fawda123/SWMPr}.} by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

<<eval = F, message = F>>=
install.packages('SWMPr')
library(SWMPr)
@
<<eval = T, echo = F, message = F, cache = F>>=
devtools::load_all('M:/docs/SWMPr')
@

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes the package structure, beginning with the retrieval functions, a description of the \texttt{swmpr} object returned after retrieval, and, finally, the organizing and analyzing functions.

@@

descrips <- c(
  'Retrieve up to 100 records\\textsuperscript{\\textit{a}} starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
  'Retrieve up to 1000 records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
  'Import site data from an independent server. These files have been downloaded from \\ac{CDMO} up to December 2014, processed using SWMPr functions, and uploaded to an Amazon server for quicker import into R.',
  'Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
  )

to_tab <- data.frame(Functions = funcs, Description = descrips, stringsAsFactors = F)
foot <- '\\textsuperscript{\\textit{a}} \\footnotesize Server limits prevent retrieving more records in a single request.'

@@

latex(
  to_tab[, 'Description', drop = F],
  file = '',
  caption = 'Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).',
  rowlabel = 'Function',
  colheads = 'Description',
  rowname = to_tab$Functions,
  caption.loc = 'top',
  col.just=c("p{3.5in}"), 
  insert.bottom = foot,
  label = 'tab:retrieve', 
  table.env = FALSE
  )
@

@@

\subsection*{The \texttt{swmpr} object class}

All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.  The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).  Correspondingg \ac{QAQC} columns for each parameter are also returned if provided by the initial data request.  The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).  Attributes of a \texttt{swmpr} object can be viewed as follows.

@@

The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for a generic function are developed for the object class.  A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods also apply to \texttt{swmpr} objects.  Available methods for the \texttt{swmpr} class are described below and can also be viewed:

@@

The ability of management and research programs to address critical environmental issues is highly dependent on the quality of data used to inform decision making.  Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessments of site-specific characteristics and more informed decisions to manage environmental resources.  The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 140 stations within the 28 estuaries of the \acl{NERRS}.  This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.  Although the data provided by \ac{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \ac{NERRS} researchers and staff to more effectively evaluate \ac{SWMP} data could be greatly improved using the SWMPr package.
